### 2025-07-17 paper/sections/intro.qmd
a very straigh-to-the-point intro now

@@

Today's prominence of artificial intelligence (AI) has largely been driven by **representation learning**: instead of relying on features and rules hand-crafted by humans, modern machine learning (ML) models are tasked with learning representations directly from the data, guided by narrow objectives such as predictive accuracy [@goodfellow2016deep]. Advances in computing have made it possible to provide these models with ever-growing degrees of freedom to achieve this task, which often allows them to outperform traditionally parsimonious models. Unfortunately, in doing so, models learn increasingly complex, sensitive representations that humans can no longer easily interpret.

The trend towards complexity for the sake of performance has come under scrutiny in recent years. At the very cusp of the deep learning (DL) revolution, @szegedy2013intriguing showed that artificial neural networks (ANN) are susceptible to adversarial examples (AEs): perturbed versions of data instances that yield vastly different model predictions despite being semantically indistinguishable from their factual counterparts. Some partial mitigation strategies have been proposed---most notably **adversarial training** [@goodfellow2014explaining]---but truly robust deep learning remains unattainable even for models that are considered "shallow" by today's standards [@kolter2023keynote].

Part of the problem is that the high degrees of freedom---high number of parameters estimated from data---provide room for many solutions that are locally optimal with respect to narrow objectives [@wilson2020case]. As one example, research on the "lottery ticket hypothesis" suggests that modern neural networks can be pruned by up to 90% without losing predictive performance [@frankle2018lottery]. Thus, looking at the predictive performance alone, found solutions may seem to provide compelling explanations for the data, when in fact they are based on purely associative and semantically meaningless patterns. This poses two related challenges. Firstly, there is no dependable way to verify if learned representations correspond to meaningful, plausible explanations. Secondly, even if we resolve this challenge, it remains undecided how to ensure that machine learning models can *only* learn valuable explanations.

The first challenge has attracted an abundance of work on **explainable AI** (XAI), a paradigm that focuses on the development of tools to derive (post-hoc) explanations from complex model representations, aiming to mitigate scenarios in which practitioners deploy opaque models and have to blindly rely on their predictions. On many occasions, this has happened in practice, causing harms to people who were adversely and unfairly affected by automated decision-making (ADM) systems involving opaque models; see, e.g., @oneil2016weapons. Effective XAI tools can also aid in monitoring models and providing recourse, empowering people to turn negative outcomes (e.g., "loan application rejected") into positive ones (e.g., "loan application accepted"). In line with this, our work builds upon **counterfactual explanations** (CE) proposed by @wachter2017counterfactual; CEs prescribe minimal changes for factual inputs that, if implemented, would prompt some fitted model to produce an alternative, more desirable output.

To our surprise, the second challenge has not yet attracted major research interest. In particular, there has been no concerted effort towards improving the degree to which learned representations promote explanations that are both **interpretable** to and deemed **plausible** by humans. Instead, the typical choice has been to improve the ability of XAI tools to identify the subset of explanations that are plausible and valid for any given model, independent of whether these explanations are compatible with the learned representations [@altmeyer2024faithful]. Fortunately, recent findings indicate that improved "explanatory capacity" of a model can arise as a consequence of regularization techniques aimed at other training objectives such as generative capacity, generalization, or robustness [@altmeyer2024faithful; @augustin2020adversarial; @schut2021generating]. Our contribution consolidates these findings within a unified framework.

Specifically, **we propose Counterfactual Training (CT)**: a novel training regime explicitly geared towards improving the explanatory capacity of models that, in high-level terms, we define as the extent to which valid counterfactual explanations derived for a model can be deemed plausible with respect to the underlying data and global actionability constraints (we refine this notion in Def. \ref{def-explainability}). To the best of our knowledge, Counterfactual Training represents the first attempt to achieve more explainable models by employing counterfactual explanations already in the training phase.

>>>>

---
bibliography: bibliography.bib
---

@@

Counterfactual explanations (CE) have become an integral part of the broader explainable AI (XAI) toolkit [@molnar2022interpretable]. Originally proposed by @wachter2017counterfactual, CEs prescribe minimal changes for factual inputs that, if implemented, would prompt some fitted model to produce an alternative, more desirable output. This is useful and necessary to not only understand how opaque AI models make predictions, but also to provide recourse to individuals subjected to them: a retail bank, for example, could use CE to provide meaningful feedback to unsuccessful loan applicants that were rejected based on an opaque AI decision-making system (ADM).

For feedback to be meaningful, counterfactual explanations need to fulfill certain desiderata [@verma2020counterfactual; @karimi2020survey]: they should be faithful to the model [@altmeyer2024faithful], plausible [@joshi2019realistic] and actionable [@ustun2019actionable]. Plausibility is typically understood as counterfactuals being *in-domain*: unsuccessful loan applicants that implement recourse should end up with a credit profile that is genuinely similar to that of individuals who have successfully repaid their loans in the past. Actionable explanations comply with practical constraints: a young, unsuccessful loan applicant cannot increase their age in an instance.

Existing state-of-the-art (SOTA) approaches in the field have largely focused on designing model-agnostic CE methods that identify subsets of counterfactuals, which comply with specific desiderata. This is problematic, because a narrow focus on any specific desideratum can adversely affect others: it is possible, for example, to generate plausible counterfactuals for models that are also highly vulnerable to implausible, possibly adversarial counterfactuals [@altmeyer2024faithful]. In this work, we therefore follow the paradigm that models---as opposed to explanation methods---should be held accountable for plausible and actionable explanations. While previous work has shown that at least plausibility can be indirectly achieved through existing techniques aimed at models' generative capacity, generalization and robustness [@altmeyer2024faithful; @augustin2020adversarial; @schut2021generating], we directly incorporate both plausibility and actionability in the training objective of models.

Specifically, **we propose Counterfactual Training (CT)**: a novel training regime for differentiable models that leverage counterfactual explanations on-the-fly to ensure that models learn plausible and actionable explanations for the data. In essence, we achieve this by contrasting faithful explanations with ground-truth data, while at the same time protecting features that are subject to actionability constraints. To the best of our knowledge, CT represents the first attempt in this direction.

In our experiments, we show that thanks to counterfactual training, (1) implausibility decreases with 16% on average; (2) the cost of reaching counterfactuals with protected features decreases with 15%; and (3) robust accuracy increases for all datasets with up to .42.

__________________________________________________________
